Synthesizing Symmetric Lenses

### 2018-07-09 papers/SymmetricLenses/main.tex
Fexed some spacing issues in Section 4

\PassOptionsToPackage{usenames,dvipsnames,svgnames,table}{xcolor}
\documentclass[acmsmall,screen,anonymous]{acmart}
\settopmatter{}

@@

\newtheorem*{theorem*}{Theorem}

@@

  \renewcommand\footnotetextcopyrightpermission[1]{} % removes footnote with conference information in first column

%\setcopyright{rightsretained}
%\acmPrice{}
%\acmDOI{10.1145/3158089}
%\acmYear{2019}
%\copyrightyear{2019}
%\acmJournal{PACMPL}
%\acmVolume{3}
%\acmNumber{POPL}
%\acmArticle{1}
%\acmMonth{1}
%\startPage{1}

@@

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011050.10011017</concept_id>
<concept_desc>Software and its engineering~Domain specific languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011066.10011070</concept_id>
<concept_desc>Software and its engineering~Application specific development environments</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Domain specific languages}
\ccsdesc[300]{Software and its engineering~Application specific development environments}

@@

\keywords{Bidirectional Programming, Program Synthesis, Type-Directed Synthesis,
Type Systems}

@@

\begin{center}
  \begin{tabular}{rcl}
    \Regex{},\RegexAlt{}
    & \GEq{} & $\String$ \\
    & \GBar{} & $\emptyset$ \\
    & \GBar{} & $\PRegexStar{\Regex}{p}$ \\
    & \GBar{} & $\RegexConcat{\Regex_1}{\Regex_2}$ \\
    & \GBar{} & $\PRegexOr{\Regex_1}{\Regex_2}{p}$
  \end{tabular}
\end{center}

@@

% begin ssl
\section{Simple Symmetric String Lenses}
\label{sec:ssl}
We now give a formal description of all the combinators in our symmetric lens
language. These lens terms are typed between pairs of stochastic regular
expressions: the typing judgment $\Lens \OfType \BRegex \Leftrightarrow
\BRegexAlt$ means that $\Lens$ is a well typed lens between
$\LanguageOf{\BRegex}$ and $\LanguageOf{\BRegexAlt}$. While we are performing
synthesis on stochastic regular expressions, lenses are typed between regular
expressions.  We can recover a regular expression from a stochastic regular
expression by deleting probability annotations.  In a slight abuse of notation,
if $\Regex$ is a stochastic regular expression, $\BRegex$ is the regular
expression obtained by removing probability annotations.

\begin{center}
  \begin{tabular}{l@{\ }l@{\ }l@{\ }l}
    % REGEX
    \Lens{} & \GEq{} & & $\IdentityLensOf{\BRegex}$\\
            & & \GBar{} & $\DisconnectOf{\BRegex}{\BRegexAlt}{\String}{\StringAlt}$ \\
            & & \GBar{} & $\IterateLensOf{\Lens}$ \\
            & & \GBar{} & $\ConcatLensOf{\Lens_1}{\Lens_2}$ \\
            & & \GBar{} & $\SwapLensOf{\Lens_1}{\Lens_2}$ \\
            & & \GBar{} & $\OrLensOf{\Lens_1}{\Lens_2}$ \\
            & & \GBar{} & $\ComposeLensOf{\Lens_1}{\Lens_2}$ \\
            & & \GBar{} & $\MergeROf{\Lens_1}{\Lens_2}$ \\
            & & \GBar{} & $\MergeLOf{\Lens_1}{\Lens_2}$ \\
            & & \GBar{} & $\InvertOf{\Lens}$ \\
  \end{tabular}
\end{center}

@@

We now provide the typing derivations and semantics for these lenses. For
simplicity of presentation, we define the typing derivations and semantics only
over unambiguous regular expressions, even though some lenses can be defined
over ambiguous regular expressions (for example, the identity lens doesn't need
any ambiguity constraints). However, we believe relaxing this restriction is
unproblematic.

\[
  \centering
  \inferrule*
  {
  }
  {
    \IdentityLensOf{\BRegex} \OfType \BRegex \Leftrightarrow \BRegex
  }
\]
\begin{center}
  \begin{tabular}{@{}r@{\ }c@{\ }l@{}}
    $\CreateR{} \App s$ & = & $s$\\
    $\CreateL{} \App s$ & = & $s$\\
    $\PutR{} \App s_1 \App s_2$ & = & $s_1$\\
    $\PutL{} \App s_1 \App s_2$ & = & $s_1$
  \end{tabular}
\end{center}

Note that the identity lens ignores the second argument in the put functions.
Because the two formats are fully synchronized, no knowledge of the prior data
is needed.

\[
  \centering
  \inferrule*
  {
    \String \in \LanguageOf{\BRegex}\\
    \StringAlt \in \LanguageOf{\BRegexAlt}
  }
  {
    \DisconnectOf{\BRegex}{\BRegexAlt}{\String}{\StringAlt}
    \OfType \BRegex \Leftrightarrow \BRegexAlt
  }
\]
\begin{center}
  \begin{tabular}{@{}r@{\ }c@{\ }l@{}}
    $\CreateR{} \App \String'$ & = & $\StringAlt$\\
    $\CreateL{} \App \StringAlt'''''$ & = & $\String$\\
    $\PutR{} \App \String''' \App \StringAlt'$ & = & $\StringAlt'$\\
    $\PutL{} \App \StringAlt' \App \String'$ & = & $\String'$
  \end{tabular}
\end{center}
Just as the identity lens ignores the second argument in puts, disconnect lenses
ignore the first in both puts and creates.  The data is unsynchronized in these
two formats, information from one format doesn't impact the other.

@@

\[
  \centering
  \inferrule*
  {
    \Lens_1 \OfType \BRegex_1 \Leftrightarrow \BRegexAlt_1\\
    \Lens_2 \OfType \BRegex_2 \Leftrightarrow \BRegexAlt_2
  }
  {
    \OrLensOf{\Lens_1}{\Lens_2} \OfType
    \RegexOr{\BRegex_1}{\BRegex_2}
    \Leftrightarrow
    \RegexOr{\BRegexAlt_1}{\BRegexAlt_2}
  }
\]
\begin{center}
  \begin{tabular}{@{}r@{\ }c@{\ }l@{}}
    $\CreateR{} \App \String$
    & =
    & $\begin{cases*}
      \Lens_1.\CreateROf{\String} & if $\String\in\LanguageOf{\BRegex_1}$\\
      \Lens_2.\CreateROf{\String} & if $\String\in\LanguageOf{\BRegex_2}$
      \end{cases*}$\\

@@

    $\PutL{} \App \StringAlt \App \String$
    & =
    & $\begin{cases*}
        \Lens_1.\PutLOf{\StringAlt}{\String} & if $\StringAlt\in\LanguageOf{\BRegexAlt_1} \BooleanAnd \String\in\LanguageOf{\BRegex_1}$\\
        \Lens_2.\PutLOf{\StringAlt}{\String} & if $\StringAlt\in\LanguageOf{\BRegexAlt_2} \BooleanAnd \String\in\LanguageOf{\BRegex_2}$\\
        \Lens_1.\CreateLOf{\StringAlt} & if $\StringAlt\in\LanguageOf{\BRegexAlt_1} \BooleanAnd \String\in\LanguageOf{\BRegex_2}$\\
        \Lens_2.\CreateLOf{\String} & if $\StringAlt\in\LanguageOf{\BRegexAlt_2} \BooleanAnd \String\in\LanguageOf{\BRegex_1}$
      \end{cases*}$\\
  \end{tabular}
\end{center}
The \OrLens lens deals with data that can come in one form or another. If the
data gets changed from one format to the other, information in the old format is
lost.

@@

\[
  \centering
  \inferrule*
  {
    \Lens_1 \OfType \BRegex_1 \Leftrightarrow \BRegexAlt\\
    \Lens_2 \OfType \BRegex_2 \Leftrightarrow \BRegexAlt
  }
  {
    \MergeROf{\Lens_1}{\Lens_2} \OfType
    \RegexOr{\BRegex_1}{\BRegex_2}
    \Leftrightarrow
    \BRegexAlt
  }
\]
\begin{center}
  \begin{tabular}{@{}r@{\ }c@{\ }l@{}}
    $\CreateR{} \App \String$
    & =
    & $\begin{cases*}
      \Lens_1.\CreateROf{\String} & if $\String\in\LanguageOf{\BRegex_1}$\\
      \Lens_2.\CreateROf{\String} & if $\String\in\LanguageOf{\BRegex_2}$
      \end{cases*}$\\

@@

    $\PutL{} \App \StringAlt \App \String$
    & =
    & $\begin{cases*}
        \Lens_1.\PutLOf{\StringAlt}{\String} & if $\String\in\LanguageOf{\BRegex_1}$\\
        \Lens_2.\PutLOf{\StringAlt}{\String} & if $\String\in\LanguageOf{\BRegex_2}$
      \end{cases*}$\\
  \end{tabular}
\end{center}
The \MergeR lens is interesting because it merges data where one data can be in
two formats, and one data has only one format. In previous work~\cite{?}, this
was combined into \OrLens{}, where \OrLens{} could have ambiguous types, but we
find it more clear to have explicit merge operators: it is easier to see what
lens the synthesis algorithm is creating.

\[
  \centering
  \inferrule*
  {
    \Lens_1 \OfType \BRegex \Leftrightarrow \BRegexAlt_1\\
    \Lens_2 \OfType \BRegex \Leftrightarrow \BRegexAlt_2
  }
  {
    \MergeLOf{\Lens_1}{\Lens_2} \OfType
    \BRegex
    \Leftrightarrow
    \RegexOr{\BRegexAlt_1}{\BRegexAlt_2}
  }
\]
\begin{center}
  \begin{tabular}{@{}r@{\ }c@{\ }l@{}}
    $\CreateR{} \App \String$
    & =
    & $\Lens_1.\CreateROf{\String}$\\

@@

    $\PutL{} \App \StringAlt \App \String$
    & =
    & $\begin{cases*}
        \Lens_1.\PutLOf{\StringAlt}{\String} & if $\StringAlt\in\LanguageOf{\BRegexAlt_1}$\\
        \Lens_2.\PutLOf{\StringAlt}{\String} & if $\StringAlt\in\LanguageOf{\BRegexAlt_2}$
      \end{cases*}$\\
  \end{tabular}
\end{center}
The \MergeL lens is symmetric to \MergeR.

\[
  \centering
  \inferrule*
  {
    \Lens_1 \OfType \BRegex \Leftrightarrow \BRegexAlt\\
    \Lens_2 \OfType \BRegexAlt \Leftrightarrow \BRegexAltAlt\\
  }
  {
    \ComposeLensOf{\Lens_1}{\Lens_2} \OfType
    \BRegex \Leftrightarrow \BRegexAltAlt
  }
\]
\begin{center}
  \begin{tabular}{@{}r@{\ }c@{\ }l@{}}
    $\CreateR{} \App \String$ & = & $\Lens_2.\CreateROf{(\Lens_1.\CreateROf{\String})}$\\
    $\CreateL{} \App \StringAlt$ & = & $\Lens_1.\CreateLOf{(\Lens_2.\CreateLOf{\StringAlt})}$\\
    $\PutR{} \App \String \App \StringAltAlt$ & = & $\Lens_2.\PutROf{(\Lens_1.\PutROf{\String}{(\Lens_2.\CreateLOf{\StringAltAlt})})}{\StringAltAlt}$\\
    $\PutL{} \App \StringAltAlt \App \String$ & = & $\Lens_1.\PutLOf{(\Lens_2.\PutLOf{\StringAltAlt}{(\Lens_2.\CreateROf{\String})})}{\String}$
  \end{tabular}
\end{center}
Composing is interesting in the put functions. Because puts require intermediary
data, we recreate that intermediary data with creates.

>>>>

\PassOptionsToPackage{usenames,dvipsnames,svgnames,table}{xcolor}
\documentclass[acmsmall,screen,anonymous]{acmart}
\settopmatter{printacmref=false}

@@

\newtheorem*{theorem*}{Theorem}
\newenvironment{centermath}
 {\begin{center}$\displaystyle}
 {$\end{center}}

@@

  \renewcommand\footnotetextcopyrightpermission[1]{} % removes footnote with conference information in first column
%\setcopyright{rightsretained}
%\acmPrice{}
%\acmDOI{10.1145/3158089}
%\acmYear{2019}
%\copyrightyear{2019}
%\acmJournal{PACMPL}
%\acmVolume{3}
%\acmNumber{POPL}
%\acmArticle{1}
%\acmMonth{1}
%\startPage{1}

@@

%\begin{CCSXML}
%<ccs2012>
%<concept>
%<concept_id>10011007.10011006.10011050.10011017</concept_id>
%<concept_desc>Software and its engineering~Domain specific languages</concept_desc>
%<concept_significance>500</concept_significance>
%</concept>
%<concept>
%<concept_id>10011007.10011006.10011066.10011070</concept_id>
%<concept_desc>Software and its engineering~Application specific development environments</concept_desc>
%<concept_significance>300</concept_significance>
%</concept>
%</ccs2012>
%\end{CCSXML}
%
%\ccsdesc[500]{Software and its engineering~Domain specific languages}
%\ccsdesc[300]{Software and its engineering~Application specific development environments}

@@

%\keywords{Bidirectional Programming, Program Synthesis, Type-Directed Synthesis,
%Type Systems}

@@

$$\Regex{},\RegexAlt{} \GEq{}  \String  \; | \; \emptyset \; | \; \PRegexStar{\Regex}{p} \; | \; \RegexConcat{\Regex_1}{\Regex_2} \; | \; \PRegexOr{\Regex_1}{\Regex_2}{p}$$

@@

\section{Simple Symmetric String Lenses}
\label{sec:ssl}
We now give a formal description of all the combinators in our symmetric lens
language. These lens terms are typed between pairs of stochastic regular
expressions: the typing judgment $\Lens \OfType \BRegex \Leftrightarrow
\BRegexAlt$ means that $\Lens$ is a well typed lens between
$\LanguageOf{\BRegex}$ and $\LanguageOf{\BRegexAlt}$. While we are performing
synthesis on stochastic regular expressions, lenses are typed between regular
expressions.  We can recover a regular expression from a stochastic regular
expression by deleting probability annotations.  In a slight abuse of notation,
if $\Regex$ is a stochastic regular expression, $\BRegex$ is the regular
expression obtained by removing probability annotations.

\begin{align*}
\ell ::= &\IdentityLensOf{\BRegex} \; | \; \DisconnectOf{\BRegex}{\BRegexAlt}{\String}{\StringAlt} \; | \; \IterateLensOf{\Lens} \; | \; \ConcatLensOf{\Lens_1}{\Lens_2} \; | \; \SwapLensOf{\Lens_1}{\Lens_2} \; | \; \OrLensOf{\Lens_1}{\Lens_2} \; | \; \ComposeLensOf{\Lens_1}{\Lens_2} \; |\\
&\MergeROf{\Lens_1}{\Lens_2} \; | \; \MergeLOf{\Lens_1}{\Lens_2} \; | \; \InvertOf{\Lens}
\end{align*}

@@

We now provide the typing derivations and semantics for these lenses. For
simplicity of presentation,\bcp{Is there any known issue with relaxing this
  restriction? I've often found ti convenient to use id lenses with ambiguous
  REs...}\afm{no known issues, I just didn't want to go into specifics about
  ambiguity in an already-full paper}\bcp{OK, we can just comment that we
  believe relaxing this restriction is unproblematic.}, we define the typing derivations and
semantics only over unambiguous regular expressions, even though some lenses can
be defined over ambiguous regular expressions (for example, the identity lens
doesn't need any ambiguity constraints).

\begin{centermath}
\begin{array}[b]{l@{\qquad \; \qquad}l}
\begin{array}[b]{c}
\quad\\
  \inferrule*
  {
  }
  {
    \IdentityLensOf{\BRegex} \OfType \BRegex \Leftrightarrow \BRegex
  }
  \\
  \quad
  \end{array}
&
\begin{array}[b]{r@{\ }c@{\ }l}
    \CreateR{} \App s & = & s\\
    \CreateL{} \App s & = & s\\
    \PutR{} \App s_1 \App s_2 & = & s_1\\
    \PutL{} \App s_1 \App s_2 & = & s_1
  \end{array}
\end{array}
\end{centermath}
Note that the identity lens ignores the second argument in the put functions.
Because the two formats are fully synchronized, no knowledge of the prior data
is needed.

\begin{centermath}
\begin{array}[b]{l@{\qquad \; \qquad}l}
  \inferrule*
  {
    \String \in \LanguageOf{\BRegex}\\
    \StringAlt \in \LanguageOf{\BRegexAlt}
  }
  {
    \DisconnectOf{\BRegex}{\BRegexAlt}{\String}{\StringAlt}
    \OfType \BRegex \Leftrightarrow \BRegexAlt
  }
&
  \begin{array}{@{}r@{\ }c@{\ }l@{}}
    $\CreateR{} \App \String'$ & = & $\StringAlt$\\
    $\CreateL{} \App \StringAlt'''''$ & = & $\String$\\
    $\PutR{} \App \String''' \App \StringAlt'$ & = & $\StringAlt'$\\
    $\PutL{} \App \StringAlt' \App \String'$ & = & $\String'$
  \end{array}
\end{array}
\end{centermath}

Just as the identity lens ignores the second argument in puts, disconnect lenses
ignore the first in both puts and creates.  The data is unsynchronized in these
two formats, information from one format doesn't impact the other.

@@

\begin{centermath}
\begin{tabular}[b]{l@{\qquad}l}
$
  \inferrule*
  {
    \Lens_1 \OfType \BRegex_1 \Leftrightarrow \BRegexAlt_1\\
    \Lens_2 \OfType \BRegex_2 \Leftrightarrow \BRegexAlt_2
  }
  {
    \OrLensOf{\Lens_1}{\Lens_2} \OfType
    \RegexOr{\BRegex_1}{\BRegex_2}
    \Leftrightarrow
    \RegexOr{\BRegexAlt_1}{\BRegexAlt_2}
  }
  $
&
  \begin{tabular}{@{}r@{\ }c@{\ }l@{}}
    $\CreateR{} \App \String$
    & =
    & $\begin{cases*}
      \Lens_1.\CreateROf{\String} & if $\String\in\LanguageOf{\BRegex_1}$\\
      \Lens_2.\CreateROf{\String} & if $\String\in\LanguageOf{\BRegex_2}$
      \end{cases*}$\\

@@

    $\PutL{} \App \StringAlt \App \String$
    & =
    & $\begin{cases*}
        \Lens_1.\PutLOf{\StringAlt}{\String} & if $\StringAlt\in\LanguageOf{\BRegexAlt_1} \BooleanAnd \String\in\LanguageOf{\BRegex_1}$\\
        \Lens_2.\PutLOf{\StringAlt}{\String} & if $\StringAlt\in\LanguageOf{\BRegexAlt_2} \BooleanAnd \String\in\LanguageOf{\BRegex_2}$\\
        \Lens_1.\CreateLOf{\StringAlt} & if $\StringAlt\in\LanguageOf{\BRegexAlt_1} \BooleanAnd \String\in\LanguageOf{\BRegex_2}$\\
        \Lens_2.\CreateLOf{\String} & if $\StringAlt\in\LanguageOf{\BRegexAlt_2} \BooleanAnd \String\in\LanguageOf{\BRegex_1}$
      \end{cases*}$\\
  \end{tabular}
\end{tabular}
\end{centermath}
The \OrLens lens deals with data that can come in one form or another. If the
data gets changed from one format to the other, information in the old format is
lost.

@@

\begin{centermath}
\begin{tabular}[b]{l@{\qquad}l}
$
  \centering
  \inferrule*
  {
    \Lens_1 \OfType \BRegex_1 \Leftrightarrow \BRegexAlt\\
    \Lens_2 \OfType \BRegex_2 \Leftrightarrow \BRegexAlt
  }
  {
    \MergeROf{\Lens_1}{\Lens_2} \OfType
    \RegexOr{\BRegex_1}{\BRegex_2}
    \Leftrightarrow
    \BRegexAlt
  }
$
&
  \begin{tabular}{@{}r@{\ }c@{\ }l@{}}
    $\CreateR{} \App \String$
    & =
    & $\begin{cases*}
      \Lens_1.\CreateROf{\String} & if $\String\in\LanguageOf{\BRegex_1}$\\
      \Lens_2.\CreateROf{\String} & if $\String\in\LanguageOf{\BRegex_2}$
      \end{cases*}$\\

@@

    $\PutL{} \App \StringAlt \App \String$
    & =
    & $\begin{cases*}
        \Lens_1.\PutLOf{\StringAlt}{\String} & if $\String\in\LanguageOf{\BRegex_1}$\\
        \Lens_2.\PutLOf{\StringAlt}{\String} & if $\String\in\LanguageOf{\BRegex_2}$
      \end{cases*}$\\
  \end{tabular}
\end{tabular}
\end{centermath}

The \MergeR lens is interesting because it merges data where one data can be in
two formats, and one data has only one format. In previous work~\cite{?}, this
was combined into \OrLens{}, where \OrLens{} could have ambiguous types, but we
find it more clear to have explicit merge operators: it is easier to see what
lens the synthesis algorithm is creating.

\begin{centermath}
\begin{tabular}[b]{l@{\qquad}l}
$
  \inferrule*
  {
    \Lens_1 \OfType \BRegex \Leftrightarrow \BRegexAlt_1\\
    \Lens_2 \OfType \BRegex \Leftrightarrow \BRegexAlt_2
  }
  {
    \MergeLOf{\Lens_1}{\Lens_2} \OfType
    \BRegex
    \Leftrightarrow
    \RegexOr{\BRegexAlt_1}{\BRegexAlt_2}
  }
$
&
  \begin{tabular}{@{}r@{\ }c@{\ }l@{}}
    $\CreateR{} \App \String$
    & =
    & $\Lens_1.\CreateROf{\String}$\\

@@

    $\PutL{} \App \StringAlt \App \String$
    & =
    & $\begin{cases*}
        \Lens_1.\PutLOf{\StringAlt}{\String} & if $\StringAlt\in\LanguageOf{\BRegexAlt_1}$\\
        \Lens_2.\PutLOf{\StringAlt}{\String} & if $\StringAlt\in\LanguageOf{\BRegexAlt_2}$
      \end{cases*}$\\
  \end{tabular}
\end{tabular}
\end{centermath}
The \MergeL lens is symmetric to \MergeR.

\begin{centermath}
\begin{tabular}[b]{l@{\qquad}l}
  $
  \inferrule*
  {
    \Lens_1 \OfType \BRegex \Leftrightarrow \BRegexAlt\\
    \Lens_2 \OfType \BRegexAlt \Leftrightarrow \BRegexAltAlt\\
  }
  {
    \ComposeLensOf{\Lens_1}{\Lens_2} \OfType
    \BRegex \Leftrightarrow \BRegexAltAlt
  }
$
&
  \begin{tabular}{@{}r@{\ }c@{\ }l@{}}
    $\CreateR{} \App \String$ & = & $\Lens_2.\CreateROf{(\Lens_1.\CreateROf{\String})}$\\
    $\CreateL{} \App \StringAlt$ & = & $\Lens_1.\CreateLOf{(\Lens_2.\CreateLOf{\StringAlt})}$\\
    $\PutR{} \App \String \App \StringAltAlt$ & = & $\Lens_2.\PutROf{(\Lens_1.\PutROf{\String}{(\Lens_2.\CreateLOf{\StringAltAlt})})}{\StringAltAlt}$\\
    $\PutL{} \App \StringAltAlt \App \String$ & = & $\Lens_1.\PutLOf{(\Lens_2.\PutLOf{\StringAltAlt}{(\Lens_2.\CreateROf{\String})})}{\String}$
  \end{tabular}
\end{tabular}
\end{centermath}
Composing is interesting in the put functions. Because puts require intermediary
data, we recreate that intermediary data with creates.
